
---
title: "Introduction to R"
author: "Dr. Ariel Mantzura"
date: "19/2/2024"
output:
  beamer_presentation:
    theme: Darmstadt
    colortheme: orchid
    fonttheme: serif
    slide_level: 2
    includes:
  slidy_presentation: default
  ioslides_presentation: default
subtitle: Data Science Course

---





## Topics to be Covered in this Lecture

- Understanding the user interface

- Installing and Loading packages

- Getting help in R

- Basic coding

- Exercises



## Installing and Loading packages

 - There is a vast amount of functions in R that can be used. 
 
 - Some of these functions are uploaded automatically every time we start working with R.
 
 - We refer to all these functions as \textbf{base R} functions.
 
 - Other functions are packed together into packages.
 
 - Typically, packages include functions that are associated with a similar task type.
 
 - Packages also include different data sets and help files.
 
 - These packages are written by a large community of \textbf{package writers}. 
 
## Installing and Loading packages

 - In order to use an R package you need to install it once on your computer.
 - Then you need to upload the package for every time you open R, i.e. the current session
 - Installing a package is with the command: 
 
 
   \textbf{\texttt{install.packages("package name")}}
   
 - Often a package depends on functions included in other packages called "dependencies".
 
 - These dependencies are installed automatically with the installation of the dependent package.
   
 - Uploading the package is with the command:
 
   \textbf{\texttt{library("package name")}}


## Installing and Loading packages

 - Multiple packages can be installed in a single command for example:
 
   \textbf{\texttt{install.packages("ggplot2","tidyverse","dplyr")}}


## Understanding the user interface

<!-- \includegraphics[width=1\textwidth,height=2.5 in]{interface.jpg} -->


## Understanding the user interface

 - The RStudio interface has 4 main panels:
 
 1. Console
 
 2. Script editor
 
 3. Environment/History
 
 4. Files/Plots/Packages/Help

## Understanding the user interface

 - The \textbf{console window} is the place where R is waiting for you to tell it what to do, and where it will show the results of a command. You can type commands directly into the console, but they will be forgotten when you close the session.
 
 - The \textbf{Rstudio script editor} allows you to ‘send’ the current line or the currently highlighted text to the R console by clicking on the Run button in the upper-right hand corner of the script editor. 
 - Alternatively, you can run by simply pressing the Ctrl and Enter keys at the same time as a shortcut.


## Understanding the user interface

 - \textbf{Environment/History}: \textbf{environment} shows all active objects and history keeps track of all commands run in console
 
 - \textbf{Files/Plots/Packages/Packages/Viewer window} 
    - \textbf{Files} lists all files in the root directory. The root(default,active) directory appears in the upper bar. 
    -  \textbf{Plots} shows any plots that you created using R. 
    -  \textbf{Packages} are a list of packages you may import. 
     - \textbf{Help} has documentation on different data sets and functions included in installed packages. for example by typing the command \texttt{?sum} you will see a detailed explanation of the function sum. 
     
## Getting help in R

 - There are many options for getting help in R.
 
 - \textbf{Help in R}: \texttt{help()} or \texttt{?}: The help() function or the help operator  ? provide access to the documentation pages for R functions, data sets and other objects.
 - For example to get help on the lm function (for linear models) type help(lm) or alternatively, type ?lm.
 - To get Access to a package that is not currently uploaded specify also the name of the package.
 - For example rlm (robust linear models) type:
   
   \textbf{\texttt{help(rlm, package="MASS")}}
 
## R help in the internet

 - \textbf{Stack Overflow} is a website for help and discussions about programming.
 
  - From Wikipedia: \textbf{Stack Overflow} is a question and answer website for computer programmers.

<!-- \includegraphics[width=0.8\textwidth,height=2 in]{stack.png} -->

## R help in the internet

<!-- \includegraphics[width=0.8\textwidth,height=2 in]{stack2.jpg} -->


# Basic Coding in R 

## Basic Coding - print a string



\textbf{\texttt{print("string")}} 

```{r pressure, echo=TRUE}
print("hello world")
```
 
 

## Replicate a string or number:
 
  
\textbf{\texttt{replicate(number of replications,"string")}} 
  
```{r pressure1, echo=TRUE}
replicate(3,"hello world")
```

## Basic Coding -  Mathematical operators: add, subtract, multiply, divide 
 
  
```{r pressure2, echo=TRUE}
5+1
5-1
5*1
5/1
```

## Mathematical operators: order of operations

- If there are no brackets, conventional order rules are applied
  
```{r p3, echo=TRUE}
(5+1)*(4-1)
5+1*4-1
4^2+3
4^(2+3)  
```
## Mathematical operators: order of operations

- n!, n over k and Modulu

  
```{r pp3, echo=TRUE}
factorial(3)     # 3!
choose(5,2)      # 5 over 2
5%%2             # modulu (remainder ofter devision) 
```


## Other mathematical functions



```{r p4, echo=TRUE}

sqrt(16);exp(3);log(2,base=exp(1))
floor(10.5);ceiling(1.5);round(10.3678,2) 

```
## Basic Coding - Creating an object

- numerical Object

```{r p5, echo=TRUE}
a=5;b<-numeric()
print(a);print(b)
d=c(1,2,3)
d
class(d)

```

## Basic Coding - Creating an object

- Character Object

```{r p6, echo=TRUE}
a=c(3,"hello",2)
a
class(a)

```




##  Basic Coding - Creating an object

- Matrix object

```{r p7, echo=TRUE}
p <- matrix(data = c(1,2,3,4,5,6,7,8,9,10),ncol = 5)
q <- matrix(data = c("a","b","c","d"),ncol = 2)
print(p);print(q)
class(p)

```
## Basic Coding - Creating an object

- Matrix object

- To acces cell, column or row use square brackets []

```{r p8, echo=TRUE}
p[1,2]
p[1,]
p[,2]
```


## Basic Coding - Creating an object

- Array object


```{r p9, echo=TRUE}
ar=array(1:6,dim=c(2,3,2))
print(ar)
```
## Basic Coding - Creating an object

- Array object

- To acces cell, column, row or entire layer by []

```{r p10, echo=TRUE}
ar[1,2,1]
ar[,,1]

```
## Basic Coding - Creating a List


 - If you want to store elements of different types in the same vector, you’ll need a list, which you create with list():
 
```{r p21, echo=TRUE}
x1 <- list(1:4, "a", TRUE)
x1

```

## Creating a list

- It is often convenient to name the components, or children, of a list:

```{r p31, echo=TRUE}
x2 <- list(a = 1:2, b = 1:3, c = 1:4)
x2


```




##  Basic Coding  - Creating a Dataframe

- Data Frames can have different types of data inside it. 
- While the first column can be character, the second and third can be numeric or logical. 
- However, each column should have the same type of data.

```{r p11, echo=TRUE}
df=data.frame(q,p[,1:2])
df
class(df)
```

## Examine dataset

- head(first rows), names of variables and dataset dimension)

```{r p12, echo=TRUE}
library(MASS)
head(iris[-3],3)
names(iris)[1:3]
dim(iris)
```

## Dataset Summary

```{r p13, echo=TRUE}
summary(iris)[1:6,-c(3,4)]

```
## Binding to datasets by rows or columns

- Column bind


```{r p14, echo=TRUE}
a=matrix(1:6,nrow=3)
b=matrix(7:12,nrow=3)
c=cbind(a,b)
d=rbind(a,b)
print(c)

```


a and b must have same number of rows

## Binding to datasets by rows or columns

- Row bind 

```{r p15, echo=TRUE}
a=matrix(1:6,nrow=3)
b=matrix(7:12,nrow=3)

d=rbind(a,b)
print(d)

```

a,b must have same number of columns

## Logical Operators



```{r p16, echo=TRUE}
a=c(1:5)
b=c(1:3)
a>3
a%in%b
b%in%a

```
## Logical Operators



```{r p17, echo=TRUE}
a=c(1:5)
a[2]=NA; is.na(a); !is.na(a)
a>3 | a<5     # or operator
a>3 & a<5     # and operator


```
## Extracting named columns

- Extracting Sepal.Length column from Iris data

```{r p19, echo=TRUE}
names(iris)[1:3]
a=iris$Sepal.Length[1:5]
print(a)
```




## Extracting data with condition

- Extract all rows bigger than 3

```{r p18, echo=TRUE}

a=iris[iris$Sepal.Length>7.5,1:3]
print(a)
```



## Exercise

 - Install and download MASS package
 - Use the command \texttt{data(package="MASS")} to observe all the data sets in the MASS package. 
 - Define the Boston dataset as an object.
 - What is the size of the data set?
 - Plot a summary of the data set.
 - Create another object that extracts from the original data set only rows that medv is larger than its 3rd quantile. 
 - Create a vector that gets the value true if tax is smaller than its first quantile.
 - Create an object that Binds this new variable to the Boston data set.
- Plot the first 5 rows of binded data set.

